Haploid Selection, Sex Ratio Bias, and Transitions Between Sex-Determination Systems

### 2018-05-01 LaTeX/main_revised.tex
simplified beginning of results, WARNING: the detail about the charpoly should be added to the appendix

% Header and Footer with logo
\usepackage{lastpage,fancyhdr,graphicx}
\usepackage[outdir=./]{epstopdf}
\pagestyle{myheadings}
\pagestyle{fancy}
\fancyhf{}
\setlength{\headheight}{27.023pt}
\lhead{\includegraphics[width=2.0in]{PLOS-submission.eps}}
\rfoot{\thepage/\pageref{LastPage}}
\renewcommand{\footrule}{\hrule height 2pt \vspace{2mm}}
\fancyheadoffset[L]{2.25in}
\fancyfootoffset[L]{2.25in}
\lfoot{\sf PLOS}
%% Include all macros below

@@

\end{flushleft}
% Please keep the abstract below 300 words
%currently 308 words
\section*{Abstract}
Sex determination is remarkably dynamic; many taxa display shifts in the location of sex-determining loci or the evolution of entirely new sex-determining systems. 
Predominant theories for why we observe such transitions generally conclude that novel sex-determining systems are favoured by selection if they equalise the sex ratio or increase linkage with a locus that experiences different selection in males vs.\ females. 
We use population genetic models to extend these theories in two ways: 
(1) We consider the dynamics of loci very tightly linked to the ancestral sex-determining loci, e.g., within the non-recombining region of the ancestral sex chromosomes. 
Variation at such loci can favour the spread of new sex-determining systems in which the heterogametic sex changes (XY to ZW or ZW to XY) and the new sex-determining region is less closely linked (or even unlinked) to the locus under selection. %, which is \textcolor{red}{not expected from previous theory}. 
(2) We consider selection upon haploid genotypes either during gametic competition (e.g., pollen competition) or meiosis (i.e., non-Mendelian segregation), which can cause the zygotic sex ratio to become biased. 
Haploid selection can drive transitions between sex-determining systems without requiring selection to act differently in diploid males vs.\ females. 
With haploid selection, we find that transitions between male and female heterogamety can evolve where linkage with the sex-determining locus is either strengthened or weakened. 
%Furthermore, we find that the selective strengths exerted by the sex-ratio and associations with selected alleles are often equivalent. 
Furthermore, we find that the selective forces to equalize the sex ratio are equally important to selective forces that generate skewed sex ratios when accounting for the spread of new sex chromosomes.
This allows sex-ratio biases to increase or decrease with the spread of new sex chromosomes and implies that transitions between sex-determining systems cannot be simply predicted by selection to equalise the sex ratio. 
Overall, our models reveal that transitions between sex-determining systems, particularly transitions where the heterogametic sex changes, can be driven by loci in previously unpredicted genomic locations that experience selection during diploid and/or haploid phases.
%Our results predict conditions under which sex-determining systems are likely to be labile and draw connections with sex-ratio evolution.

% Please keep the Author Summary between 150 and 200 words
% Use first person. PLOS ONE authors please skip this step. 
% Author Summary not valid for PLOS ONE submissions.   
%currently 193 words
\section*{Author summary}
Systems of sex determination are strikingly diverse and labile in many clades. 
This poses the question: what drives transitions between sex-determining systems? 
Here, we use models to derive conditions under which new sex-determining systems spread. 
Prevailing views suggest that new sex-determining systems are favoured when they equalize the sex ratio and/or when they are more closely linked to genes that experience differential selection in males and females. 
Our models include selection upon haploid genotypes (meiotic drive or gametic competition), which biases the sex-ratio and occurs differently in male and female gametes. 
Surprisingly, we find the two forces (selection to equalize the sex ratio and the benefits of hitchhiking with driven alleles that distort the sex ratio) will often be equally strong, and thus neither is sufficient to explain the spread of new sex-determining systems in every case. 
We also find that new sex-determining alleles can spread despite being less closely linked to selected loci, driven by initially tight linkage or haploid selection. 
Our models therefore predict that loci in previously unexpected genomic locations and/or experiencing various types of selection (including haploid selection) can now be implicated as drivers of transitions between sex-determining systems.

@@

%Tight linkage Model Feature and Result
One novel feature of the models developed here is that we explicitly consider the maintenance of genetic variation around the ancestral sex-determining locus (e.g., within the non-recombining region of a sex chromosome). %during trans-GSD transitions
%, when considering loci in very tight linkage with the ancestral sex-determining locus (e.g., within the non-recombining region of a sex chromosome), 
%\textcolor{red}{[vDK2007 does "explicitly" consider this as well (although on pg 15 of their SOM they say they will ignore equilibrium A, as it is stable under a "limited range of biologically relevant parameter combinations"; perhaps haploid selection changes this impression), but only for cis-GSD transitions. They find that if equil B is stable the neo-Y never spreads when r=0. vDK2010 does not "explicitly" calculate pX and pY, which are required to see if their equation 7 is positive or negative, but nowhere do they seem to say that weaker linkage \textit{can't} evolve given r initially small.]}
%we explicitly calculate ancestral polymorphisms. 
%Our analysis reveals that, under various forms of selection, the ancestral-X can carry suboptimal alleles for females and/or the ancestral-Y can carry female-beneficial alleles. 
Counterintuitively, when linkage is tight between the sex-determining locus and a selected locus, an allele good for females can be at higher frequency on the ancestral-Y than on the ancestral-X under a variety of forms of selection.
This, in turn, can favour a new ZW sex-determining locus that has weaker linkage with loci under selection (a similar argument applies to ZW$\rightarrow$XY transitions), which was not \textcolor{red}{apparent/discovered/considered} in previous theory \cite{vanDoorn:2010hu}. 
That is, we show that selected loci in very tight linkage with the ancestral GSD locus can favour trans-GSD transitions during which linkage associations are actually weakened.

%Haploid selection model feature and discussion of why needed
Most significantly, we include haploid selection (gametic competition or meiotic drive) in models describing cis-GSD, trans-GSD, and GSD to ESD transitions.
This poses an apparent evolutionary problem. 
On one hand, haploid selection is typically sex-specific in that it usually occurs among gametes produced by one sex only~\cite{Mulcahy:1996ha,JOSEPH:2004haa,Ubeda:2005gw,Lindholm:2016cw}. 
Therefore, one might expect new sex-determining systems to benefit from close linkage with haploid selected loci, as found for sex-differences in diploid selection~\cite{vanDoorn:2007eu,vanDoorn:2010hu, Muralidhar2018}. 
On the other hand, associations between sex-determining loci and haploid selected loci generate biased zygotic sex ratios, which should generally hinder the spread of new sex-determining systems.

%Previous models with haploid selection
Two previous studies have considered the spread of GSD with sex-specific meiotic drive~\cite{Kozielska:2010vm,Ubeda:2015fx} under a limited number of possible genetic architectures and diploid selective regimes. 
Ubeda et al. (2015)~\cite{Ubeda:2015fx} considered ancestral-ESD (with no sex-ratio bias) and numerically showed that new GSD alleles can spread if they arise in linkage with meiotic drive loci. 
For example, a masculinizing allele spreads in association with an allele that is favoured during male meiosis, causing sex ratios to become male-biased.
This suggests that the benefits of associating with driving alleles overwhelms selection to balance the sex ratio and that haploid selection should cause increased sex-linkage. %, at least when new sex-determining alleles are rare. 
However, Kozielska et al. (2010)~\cite{Kozielska:2010vm} considered an ancestral GSD system that is perfectly linked to a meiotic driver (creating a sex ratio bias). 
They found that new, completely unlinked, GSD systems can spread if they generate the rarer sex, creating a balanced sex ratio. 
This suggests that Fisherian sex-ratio selection overwhelms the benefits of being associated with driving alleles and that haploid selection should cause decreased sex-linkage.

%main results
Here, we analytically find the conditions under which new GSD or ESD systems spread in ancestral GSD systems with generic linkage between the loci involved and generic sex-specific haploid and diploid selection. 
Doing so, we reconcile the results of Kozielska et al. (2010)~\cite{Kozielska:2010vm} and Ubeda et al. (2015)~\cite{Ubeda:2015fx} by showing when new GSD systems that increase \textit{or} decrease linkage with loci under haploid selection spread. %that experience haploid selection.
This result is qualitatively distinct from those for diploid selection alone~\cite{vanDoorn:2007eu,vanDoorn:2010hu} and suggests that haploid selection is more likely to promote transitions between sex-determination systems. 
This also implies that transitions involving haploid selection cannot be simply explained by invoking sex-ratio selection. 
In particular, we show that under a wide range of conditions, transitions to sex-determining systems that increase or decrease the sex-ratio bias are favoured \textit{equally strongly}. %(and in situations where sex-ratio biases are ancestrally present or absent). 
Finally, we show that ESD may not evolve, even if the sex ratio is initially biased by haploid selection, which is not predicted by previous theories for transitions to ESD~\cite{Charnov:1982wg,Bull:1983vi,West:2009we}. 
%Together, our results suggest that neither selection to equalize the sex ratio nor the benefits of associating with haploid selected alleles is an overwhelming force driving transitions between sex-determining systems. 
Together, our results suggest that both selection to equalize the sex ratio and the benefits of associating with haploid selected alleles can drive transitions between sex-determining systems, leading to tighter or weaker sex-linkage.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%Model Outline
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}[!h]
%\centering
\centerline{\includegraphics[width=\linewidth]{../Plots/Sex_determination_outline.pdf}}
\caption{
{\bf Outline of model features.}
Panel A: Recombination rate parameters between the ancestral-sex-determining locus ($\mathbf{X}$, here assumed to have alleles X and Y), a locus under selection ($\mathbf{A}$, with alleles $A$ and $a$), and a new sex-determining locus ($\mathbf{M}$, with alleles $M$ and $m$). 
If $r<1/2$, then associations between ancestral sex-determining alleles and selected alleles can be maintained in males. 
Panel B: Haploid selection is often sex-specific, occurring during haploid production or competition in either males or females. 
For example, haploid selection in males only is represented by the dashed circle.  
If X or Y alleles remain associated with alleles that experience haploid selection in males ($r<1/2$), then zygotic sex ratios can become biased because either X- or Y-bearing male gametes/gametophytes will be more abundant after haploid selection. 
However, the zygotic sex ratio is not biased by male haploid selection in ZW sex-determining systems. 
Similarly, zygotic sex-ratio biases can arise if haploid selected alleles are associated with new sex-determining alleles ($R<1/2$). 
Panel C: During cis-GSD transitions (XY to XY or ZW to ZW), a neo-Y allele spreads to pseudo-fixation (its maximum frequency among male gametes) and the ancestral Y allele is lost. 
Panel D: During trans-GSD transitions (XY to ZW or ZW to XY), a neo-W allele spreads to pseudo-fixation (its maximum frequency among female gametes) and the ancestral X allele is lost. 
Neo-W alleles allow Y-associated alleles into females, which may impede or aid their spread. 
}
\label{fig:model_outline}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

We consider two forms of selection upon haploid genotypes, `gametic competition' and `meiotic drive'. 
During gametic competition, we assume that a representative sample of all gametes/gametophytes (hereafter gametes) compete with others of the same sex for fertilization, which implies a polygamous mating system. 
Relative fitnesses in sex $\circ \in \{\female,\male\}$ during gametic competition are given by $w_A^\circ$ and $w_a^\circ$ (see table \ref{tab:fitnesstable}).
On the other hand, meiotic drive in our model only affects the segregation of gametes produced by heterozgotes.
Specifically, gametes produced by $Aa$ heterozgotes of sex $\circ$ bear allele $A$ with probability $\alpha^\circ$. 
We note that competition between sperm produced by a single male (e.g. in a monogamous mating system) would be appropriately modelled as male meiotic drive, as only the frequency of gametes produced by heterozygotes would be affected. 
However, we do not consider scenarios in which there is competition among gametes produced by a small number of males/females (e.g.,~\cite{Holman:2015en}).

In each generation, we census the genotype frequencies in male and female gametes before gametic competition. 
After gametic competition, conjugation between male and female gametes occurs at random.
The resulting zygotes develop as males or females, depending on their genotypes at the $\mathbf{X}$ and $\mathbf{M}$ loci. %(and the $\mathbf{M}$ genotype of their mother in the case of maternal control).
Diploid males and females then experience viability and/or individual-based fertility selection, with relative fitnesses $w_{AA}^{\circ}$, $w_{Aa}^{\circ}$, and $w_{aa}^{\circ}$. %where $g$ is the diploid genotype at the $\mathbf{A}$ locus ($g \in \{AA, Aa, aa\}$).  
We do not consider fertility selection based on the mating pair or sexual selection.
The next generation of gametes is produced by meiosis, during which recombination and sex-specific meiotic drive can occur. 
Recombination (i.e., an odd number of cross-overs) occurs between loci $\mathbf{X}$ and $\mathbf{A}$ with probability $r$, between loci $\mathbf{A}$ and $\mathbf{M}$ with probability $R$, and between loci $\mathbf{X}$ and $\mathbf{M}$ with probability $\rho$.
Any linear order of the loci can be modelled with appropriate choices of $r$, $R$, and $\rho$ (see Fig \ref{fig:model_outline}A and \nameref{tab:chisubstitutions}). 
Our model is entirely deterministic and hence ignores chance fluctuations in allele frequencies due to genetic drift.
%During the , the $\mathbf{A}$ locus can experience sex-specific gametic competition, diploid selection, and/or meiotic drive.

\begin{table}[ht]
%\centering
\smallskip
\caption{{\bf Relative fitness of different genotypes in sex $\circ \in \{\female,\male\}$ } }
\begin{tabular}{l c }
\hline\hline
  Genotype & Relative fitness during gametic competition \\ [0.5ex] \hline
  $A$ & $w_{A}^\circ = 1+t^\circ$ \\
  $a$ & $w_{a}^\circ = 1$ \\ [0.5ex] \hline
  Genotype & Relative fitness during diploid selection \\ [0.5ex] \hline
  $AA$ & $w_{AA}^\circ = 1+ s^\circ$ \\
  $Aa$ & $w_{Aa}^\circ = 1+h^\circ s^\circ$ \\
  $aa$ & $w_{aa}^\circ = 1$ \\ [0.5ex] \hline
  Genotype & Transmission during meiosis in $Aa$ heterozygotes \\ [0.5ex] \hline
  $A$ & $\alpha^\circ=1/2+\alpha_{\Delta}^{\circ}/2$ \\
  $a$ & $1-\alpha^\circ=1/2-\alpha_{\Delta}^{\circ}/2$ \\
  \hline \hline
  \label{tab:fitnesstable}
 \end{tabular}
\end{table}

@@

In the ancestral population, it is convenient to follow the frequency of the $A$ allele among female gametes (eggs), $p^\female_X$, and among X-bearing, $p^\male_X$, and among Y-bearing, $p^\male_Y$, male gametes (sperm/pollen). 
We also track the fraction of male gametes that are Y-bearing, $q$, which may deviate from $1/2$ due to meiotic drive in males. 
We consider only equilibrium frequencies of alleles, $\hat{p}^\circ_i$, and Y-bearing male gametes, $\hat{q}$, when determining the invasion of new sex-determining factors.

@@

@@

We begin by describing the conditions under which new sex determining alleles can spread within a population, without explicitly specifying ancestral allele frequencies.
These general conditions then allow us to consider several special cases of interest in subsequent sections, where equilibrium ancestral allele frequencies are explicitly calculated.

The evolution of a new sex-determining system depends on the growth rate of a rare mutant allele, $m$, at the novel sex-determining locus, $\mathbf{M}$. %, increases in frequency when rare. 
This growth rate is determined by the leading eigenvalue ($\lambda_m^{(XY)}$) of the system of eight equations describing the frequency of eggs and sperm carrying the $m$ allele in the next generation (equations \ref{eq:recursions}). %, evaluated at the above equilibrium.
%Thus, the system describing the change in frequency of the $m$ allele consists of eight recursion equations. 
This system simplifies substantially for an epistatically dominant neo-Y ($k=0$) or neo-W ($k=1$), see \nameref{app:inv_cond}. 
The leading eigenvalue for a rare neo-Y or neo-W allele, $m\in\{Y',W'\}$, is the largest value of $x$ that solves $x^2+ b x + c = 0$, where $b= - (\Lambda_{mA}^{(XY)} + \Lambda_{ma}^{(XY)})+(\chi_{mA}^{(XY)} + \chi_{ma}^{(XY)})$ and $c = (\Lambda_{mA}^{(XY)} - \chi_{mA}^{(XY)}) (\Lambda_{ma}^{(XY)} - \chi_{ma}^{(XY)}) - \chi_{mA}^{(XY)} \chi_{ma}^{(XY)}$.
The variable $\Lambda_{mi}^{(XY)}>0$ is the multiplicative growth rate of the neo-sex-determining allele $m$ on background $i\in\{A,a\}$ without accounting for loss due to recombination ($R=0$).
We will call  $\Lambda_{mi}^{(XY)}$ the ``haplotypic growth rate''.
The variable $\chi_{mi}^{(XY)}>0$ is the rate at which mutant haplotypes on background $i\in\{A,a\}$ recombine onto the other $\mathbf{A}$ locus background in heterozygotes (proportional to $R$).
We call $\chi_{mi}^{(XY)}$ the ``dissociative force", as it breaks down linkage disequilibrium.
Both variables are listed in Table \ref{tab:haplotype_growth} for neo-Y and neo-W mutants.

>>>>

% Header and Footer with logo
\usepackage{lastpage,fancyhdr,graphicx}
\usepackage{epstopdf}
\pagestyle{myheadings}
\pagestyle{fancy}
\fancyhf{}
\setlength{\headheight}{27.023pt}
\lhead{\includegraphics[width=2.0in]{PLOS-submission.eps}}
\rfoot{\thepage/\pageref{LastPage}}
\renewcommand{\footrule}{\hrule height 2pt \vspace{2mm}}
\fancyheadoffset[L]{2.25in}
\fancyfootoffset[L]{2.25in}
\lfoot{\sf PLOS}
%% Include all macros below

@@

\end{flushleft}
% Please keep the abstract below 300 words
%currently 308 words
\section*{Abstract}
Sex determination is remarkably dynamic; many taxa display shifts in the location of sex-determining loci or the evolution of entirely new sex-determining systems. 
Predominant theories for why we observe such transitions generally conclude that novel sex-determining systems are favoured by selection if they equalise the sex ratio or increase linkage with a locus that experiences different selection in males vs.\ females. 
We use population genetic models to extend these theories in two ways: 
(1) We consider the dynamics of loci very tightly linked to the ancestral sex-determining loci, e.g., within the non-recombining region of the ancestral sex chromosomes. 
Variation at such loci can favour the spread of new sex-determining systems in which the heterogametic sex changes (XY to ZW or ZW to XY) and the new sex-determining region is less closely linked (or even unlinked) to the locus under selection, which is not expected from previous theory. 
(2) We consider selection upon haploid genotypes either during gametic competition (e.g., pollen competition) or meiosis (i.e., non-Mendelian segregation), which can cause the zygotic sex ratio to become biased. 
Haploid selection can drive transitions between sex-determining systems without requiring selection to act differently in diploid males vs.\ females. 
With haploid selection, we find that transitions between male and female heterogamety can evolve where linkage with the sex-determining locus is either strengthened or weakened. 
%Furthermore, we find that the selective strengths exerted by the sex-ratio and associations with selected alleles are often equivalent. 
Furthermore, we find that sex-ratio biases may increase or decrease with the spread of new sex chromosomes, which implies that transitions between sex-determining systems cannot be simply predicted by selection to equalise the sex ratio.
In fact, under many conditions, we find that transitions in sex determination are favoured equally strongly in cases where the sex ratio bias increases or decreases.
%Furthermore, we find that the selective forces to equalize the sex ratio are equally important to selective forces that generate skewed sex ratios when accounting for the spread of new sex chromosomes.
%This allows sex-ratio biases to increase or decrease with the spread of new sex chromosomes and implies that transitions between sex-determining systems cannot be simply predicted by selection to equalise the sex ratio. 
Overall, our models reveal that transitions between sex-determining systems, particularly transitions where the heterogametic sex changes, can be driven by loci in previously unpredicted genomic locations that experience selection during diploid and/or haploid phases.
%Our results predict conditions under which sex-determining systems are likely to be labile and draw connections with sex-ratio evolution.

% Please keep the Author Summary between 150 and 200 words
% Use first person. PLOS ONE authors please skip this step. 
% Author Summary not valid for PLOS ONE submissions.   
%currently 193 words
\section*{Author summary}
Systems of sex determination are strikingly diverse and labile in many clades. 
This poses the question: what drives transitions between sex-determining systems? 
Here, we use models to derive conditions under which new sex-determining systems spread. 
Prevailing views suggest that new sex-determining systems are favoured when they equalize the sex ratio and/or when they are more closely linked to genes that experience differential selection in males and females. 
Our models include selection upon haploid genotypes (meiotic drive or gametic competition), which causes sex-ratio biases and occurs differently in male and female gametes. 
Surprisingly, we find the two forces (selection to equalize the sex ratio and the benefits of hitchhiking with driven alleles that distort the sex ratio) will often be equally strong, and thus neither is sufficient to explain the spread of new sex-determining systems in every case. 
Even more unexpectedly, we find that new sex-determining alleles can spread despite being less closely linked (or unlinked) to selected loci. 
Therefore, our models predict loci in previously unexpected genomic locations and/or experiencing various types of selection (including haploid selection) can now be implicated as drivers of transitions between sex-determining systems.

@@

%Tight linkage Model Feature and Result
One novel feature of the models developed here is that we explicitly consider the maintenance of genetic variation around the ancestral sex-determining locus (e.g., within the non-recombining region of a sex chromosome). %during trans-GSD transitions
%, when considering loci in very tight linkage with the ancestral sex-determining locus (e.g., within the non-recombining region of a sex chromosome), 
%\textcolor{red}{[vDK2007 does "explicitly" consider this as well (although on pg 15 of their SOM they say they will ignore equilibrium A, as it is stable under a "limited range of biologically relevant parameter combinations"; perhaps haploid selection changes this impression), but only for cis-GSD transitions. They find that if equil B is stable the neo-Y never spreads when r=0. vDK2010 does not "explicitly" calculate pX and pY, which are required to see if their equation 7 is positive or negative, but nowhere do they seem to say that weaker linkage \textit{can't} evolve given r initially small.]}
%we explicitly calculate ancestral polymorphisms. 
%Our analysis reveals that, under various forms of selection, the ancestral-X can carry suboptimal alleles for females and/or the ancestral-Y can carry female-beneficial alleles. 
Counterintuitively, when linkage is tight between the sex-determining locus and a selected locus, an allele good for females can be at higher frequency on the ancestral-Y than on the ancestral-X under a variety of forms of selection.
This, in turn, can favour a new ZW sex-determining locus that has weaker linkage with loci under selection, which was not considered in previous theory \cite{vanDoorn:2010hu}. A similar argument applies to ZW$\rightarrow$XY transitions. 
Thus, we show that selected loci in very tight linkage with the ancestral GSD locus can favour trans-GSD transitions during which linkage associations are actually weakened.

%Haploid selection model feature and discussion of why needed
Most significantly, we include haploid selection (gametic competition or meiotic drive) in models describing cis-GSD, trans-GSD, and GSD to ESD transitions, which poses an apparent evolutionary problem. 
On one hand, haploid selection is typically sex-specific in that it usually occurs among gametes produced by one sex only~\cite{Mulcahy:1996ha,JOSEPH:2004haa,Ubeda:2005gw,Lindholm:2016cw}. 
Therefore, one might expect new sex-determining systems to benefit from close linkage with haploid selected loci, as found for sex-differences in diploid selection~\cite{vanDoorn:2007eu,vanDoorn:2010hu, Muralidhar2018}. 
On the other hand, associations between sex-determining loci and haploid selected loci generate biased zygotic sex ratios, which should generally hinder the spread of new sex-determining systems.

%Previous models with haploid selection
Two previous studies have considered the spread of GSD with sex-specific meiotic drive~\cite{Kozielska:2010vm,Ubeda:2015fx} under a limited number of possible genetic architectures and diploid selective regimes. 
Ubeda et al. ~\cite{Ubeda:2015fx} consider ancestral-ESD (with no sex-ratio bias) and numerically show that new GSD alleles can spread if they arise in linkage with meiotic drive loci. 
For example, a masculinizing allele spreads in association with an allele that is favoured during male meiosis, causing sex ratios to become male-biased and thereby suggesting that such beneficial associations overwhelm selection to balance the sex ratio. 
However, Kozielska et al. (2010)~\cite{Kozielska:2010vm} consider meiotic drive on an ancestral GSD sex chromosome (creating a sex ratio bias). 
They find that new, completely unlinked, GSD systems can spread if they generate the rarer sex, creating a balanced sex ratio. 
That is, they conclude that Fisherian sex-ratio selection caused by haploid selection favours new GSD systems.

%main results
Here, we analytically find the conditions under which new GSD or ESD systems spread with generic linkage between the loci involved (and generic sex-specific haploid and diploid selection). 
Strikingly, we reconcile the results of Kozielska et al.~\cite{Kozielska:2010vm} and Ubeda et al.~\cite{Ubeda:2015fx} by deriving conditions for the spread of new GSD systems that increase \textit{or} decrease linkage with haploid selected loci. %that experience haploid selection.
Our result is qualitatively distinct from those for diploid selection alone~\cite{vanDoorn:2007eu,vanDoorn:2010hu} and suggests that haploid selection is more likely to promote transitions between sex-determination systems. 
We also show that transitions involving haploid selection cannot be simply explained by invoking sex-ratio selection. 
In particular, under a wide range of conditions, we show that transitions in sex-determining system are favoured \textit{equally strongly} in situations where sex-ratio biases increase or decrease (and in situations where sex-ratio biases are ancestrally present or absent). 
Finally, we show that ESD may not evolve, even if the sex ratio is initially biased by haploid selection, which would not be predicted by previous theories for transitions to ESD~\cite{Charnov:1982wg,Bull:1983vi,West:2009we}. 
Together, these results suggest that selection to equalize the sex ratio is not an overwhelming force driving transitions between sex-determining systems.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%Model Outline
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}[!h]
%\centering
\centerline{\includegraphics[width=\linewidth]{../Plots/Sex_determination_outline.pdf}}
\caption{
{\bf Outline of model features.}
Panel A: Recombination rate parameters between the ancestral-sex-determining locus ($\mathbf{X}$, here assumed to have alleles X and Y), a locus under selection ($\mathbf{A}$, with alleles $A$ and $a$), and a new sex-determining locus ($\mathbf{M}$, with alleles $M$ and $m$). 
If $r<1/2$, then associations between ancestral sex-determining alleles and selected alleles can be present in new male gametes/gametophytes. 
Panel B: Haploid selection is often sex-specific, occurring during haploid production or competition in either males or females. 
For example, haploid selection in males only is represented by the dashed circle.  
If X or Y alleles remain associated with alleles that experience haploid selection in males ($r<1/2$), then zygotic sex ratios can become biased because either X- or Y-bearing male gametes/gametophytes will be more abundant after haploid selection. 
However, the zygotic sex ratio is not biased by male haploid selection in ZW sex-determining systems. 
Similarly, zygotic sex-ratio biases can arise if haploid selected alleles are associated with new sex-determining alleles ($R<1/2$). 
Panel C: During cis-GSD transitions (XY to XY or ZW to ZW), a neo-Y allele spreads to pseudo-fixation (its maximum frequency among male gametes) and the ancestral Y allele is lost. 
Panel D: During trans-GSD transitions (XY to ZW or ZW to XY), a neo-W allele spreads to pseudo-fixation (its maximum frequency among female gametes) and the ancestral X allele is lost. 
Neo-W alleles allow Y-associated alleles into females, which may impede or aid their spread. 
}
\label{fig:model_outline}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

We consider two forms of selection upon haploid genotypes, `gametic competition' and `meiotic drive'. 
During gametic competition, we assume that a representative sample of all gametes/gametophytes (hereafter gametes) compete for fertilization, which implies a polygamous mating system. 
Relative fitnesses during gametic competition are given by $w_A^\circ$ and $w_a^\circ$ ($\circ \in \{\female,\male\}$; see table \ref{tab:fitnesstable}).
On the other hand, meiotic drive in our model only affects the segregation of gametes produced by heterozgotes.
Specifically, gametes produced by $Aa$ heterozgotes of sex $\circ$ bear allele $A$ with probability $\alpha^\circ$. 
We note that competition between sperm produced by a single male (e.g. in a monogamous mating system) would be appropriately modelled as male meiotic drive, as only the frequency of gametes produced by heterozygotes would be affected. 
However, we do not consider scenarios in which there is competition among gametes produced by a small number of males/females (e.g.,~\cite{Holman:2015en}).

In each generation, we census the genotype frequencies in male and female gametes before gametic competition. 
Below, it is convenient to follow the ancestral equilibrium frequency of the $A$ allele among female gametes (eggs), $\hat{p}^\female_X$, and among X-bearing, $\hat{p}^\male_X$, and among Y-bearing, $\hat{p}^\male_Y$, male gametes (sperm/pollen). 
We also track the ancestral fraction of male gametes that are Y-bearing, $\hat{q}$, which may deviate from $1/2$ due to meiotic drive in males.

After gametic competition, conjugation between male and female gametes occurs at random.
The resulting zygotes develop as males or females, depending on their genotypes at the $\mathbf{X}$ and $\mathbf{M}$ loci. %(and the $\mathbf{M}$ genotype of their mother in the case of maternal control).
Diploid males and females then experience viability and/or individual-based fertility selection, with relative fitnesses $w_{AA}^{\circ}$, $w_{Aa}^{\circ}$, and $w_{aa}^{\circ}$. %where $g$ is the diploid genotype at the $\mathbf{A}$ locus ($g \in \{AA, Aa, aa\}$).  
We do not consider fertility selection that depends on the mating partner, e.g., sexual selection with variation in choosiness. 
The next generation of gametes is produced by meiosis, during which recombination and sex-specific meiotic drive can occur. 
Recombination (i.e., an odd number of cross-overs) occurs between loci $\mathbf{X}$ and $\mathbf{A}$ with probability $r$, between loci $\mathbf{A}$ and $\mathbf{M}$ with probability $R$, and between loci $\mathbf{X}$ and $\mathbf{M}$ with probability $\rho$.
Any linear order of the loci can be modelled with appropriate choices of $r$, $R$, and $\rho$ (see Fig \ref{fig:model_outline}A and \nameref{tab:chisubstitutions}). 
Our model is entirely deterministic and hence ignores chance fluctuations in allele frequencies due to genetic drift.
%During the , the $\mathbf{A}$ locus can experience sex-specific gametic competition, diploid selection, and/or meiotic drive.

\begin{table}[ht]
%\centering
\smallskip
\caption{{\bf Relative fitness of different genotypes in sex, $\circ \in \{\female,\male\}$ } }
\begin{tabular}{l c }
\hline\hline
  Genotype & Relative fitness during gametic competition \\ [0.5ex] \hline
  $A$ & $w_{A}^\circ = 1+t^\circ$ \\
  $a$ & $w_{a}^\circ = 1$ \\ [0.5ex] \hline
  Genotype & Relative fitness during diploid selection \\ [0.5ex] \hline
  $AA$ & $w_{AA}^\circ = 1+ s^\circ$ \\
  $Aa$ & $w_{Aa}^\circ = 1+h^\circ s^\circ$ \\
  $aa$ & $w_{aa}^\circ = 1$ \\ [0.5ex] \hline
  Genotype & Transmission during meiosis in $Aa$ heterozygotes \\ [0.5ex] \hline
  $A$ & $\alpha^\circ=1/2+\alpha_{\Delta}^{\circ}/2$ \\
  $a$ & $1-\alpha^\circ=1/2-\alpha_{\Delta}^{\circ}/2$ \\
  \hline \hline
  \label{tab:fitnesstable}
 \end{tabular}
\end{table}

@@

@@

We begin by describing the general conditions under which new genetic sex determining alleles can spread within a population.  
These general conditions then allow us to consider several special cases of interest in subsequent sections. 
Finally, we consider the spread of alleles that specify environmental sex determination.

@@

The evolution of a new sex-determining system requires that a rare mutant allele, $m$, at the novel sex-determining locus, $\mathbf{M}$, increases in frequency when rare (i.e., $\lambda_m^{(XY)}>1$, where $\lambda_m^{(XY)}$ is the leading eigenvalue of the system of eight equations describing $m$-bearing gamete frequencies, \ref{eq:recursions}). 
%This is determined by the leading eigenvalue ($\lambda_m^{(XY)}$) of the system of eight equations describing the frequency of eggs and sperm carrying the $m$ allele in the next generation (equations \ref{eq:recursions}). 
This system simplifies substantially for an epistatically dominant neo-Y ($k=0$) or neo-W ($k=1$), see \nameref{app:inv_cond}. 
%We will consider only equilibrium frequencies of alleles, $\hat{p}^\circ_i$, and Y-bearing male gametes, $\hat{q}$, when calculating the eigenvalues.

Invasion by a neo-Y or a neo-W primarily depends on the ``haplotypic growth rates'' (denoted by $\Lambda_{mi}^{(XY)}$) of the neo-sex determination allele $m$ on background $i\in\{A,a\}$ without accounting for loss due to recombination ($R=0$), see Table \ref{tab:haplotype_growth}.
If both haplotypic growth rates are greater than one ($\Lambda_{mA}^{(XY)}, \Lambda_{ma}^{(XY)}>1$), then the new sex-determining allele invades regardless of the rate of recombination between the new sex-determining locus and the selected locus ($R$), see \nameref{app:inv_cond} for details.
Conversely, if both haplotypic growth rates are less than one ($\Lambda_{mA}^{(XY)}, \Lambda_{ma}^{(XY)}<1$), then invasion can never occur.  
Finally, if only one haplotypic growth rate is greater than one, the new sex-determining allele can always invade when arising at a locus that is tightly linked to the selected locus ($R\approx0$).  
Furthermore, it can be shown that the leading eigenvalue declines with $R$, and invasion requires that $R$ is sufficiently small such that: 
\begin{equation}\label{eq:lambdasGen}
%\frac{\chi_{mA}}{\Lambda_{mA}-1}+\frac{\chi_{ma}}{\Lambda_{ma}-1}>0.
\chi_{ma}^{(XY)}/\left(\Lambda_{ma}^{(XY)} - 1\right) + \chi_{mA}^{(XY)}/\left(\Lambda_{mA}^{(XY)} - 1\right) < 1
\end{equation}
\noindent 
where $\chi_{mi}^{(XY)}>0$ is the rate at which mutant haplotypes on background $i\in\{A,a\}$ recombine onto the other $\mathbf{A}$ locus background in heterozygotes (proportional to $R$; we call this the ``dissociative force" as it breaks down linkage disequilibrium), see Table \ref{tab:haplotype_growth}.
Condition \ref{eq:lambdasGen} may or may not be satisfied for the full range of locations of the new sex-determining locus, including $R=1/2$, depending on the nature of selection.  
Interpreting this condition, if we assume that only the $mA$ haplotype would increase in frequency when $R=0$ ($\Lambda_{ma}^{(XY)}<1<\Lambda_{mA}^{(XY)}$) then the first term on the left-hand side of \eqref{eq:lambdasGen} is negative and invasion requires that growth rate of $mA$ haplotypes ($\Lambda_{mA}^{(i)}-1>0$) and the rate at which they are produced by recombination ($\chi_{ma}^{(i)}$) are sufficiently large relative to the rate of decline of $ma$ haplotypes ($1-\Lambda_{ma}^{(i)}>0$) and the rate at which $m$ and $A$ are dissociated by recombination ($\chi_{mA}^{(i)}$).
